---
title: "Class 12 pt. 2: Population analysis"
author: Morgan Black (PID A14904860)
format: pdf
---

## Class 12 Section 4: Population scale analysis

```{r}

expr <- read.table("rs8067378_ENSG00000172057.6.txt")
nrow(expr)

```

### Q13: Determine the sample size for each genotype and their corresponding median expression levels for each of these genotypes
```{r}

#How many data points are there for each genotype?
table(expr$geno)

```

```{r}

#What's the median expression level for each of these genotypes?

#For the A/A genotype:
AA_geno <- subset(expr, geno == "A/A")
median(AA_geno$exp)

#For the A/G genotype:
AG_geno <- subset(expr, geno == "A/G")
median(AG_geno$exp)

#For the G/G genotype:
GG_geno <- subset(expr, geno == "G/G")
median(GG_geno$exp)

```



### Q14: Generate a boxplot with a box per genotype, what could you infer from the relative expression value between A/A and G/G displayed in this plot? Does the SNP effect the expression of ORMDL3?
```{r}

library(ggplot2)
ggplot(expr) + aes(geno, exp, fill=geno) +
  geom_boxplot(notch=TRUE)


```
The relative expression levels are much lower in the G/G genotype than in the A/A genotype, showing that the G/G SNP clearly has an effect on ORMLD3 gene expression by leading to much lower expression in this population. 
