---
title: "Class 9: Halloween Candy Mini-Project"
author: "Morgan Black (PID A14904860)"
format: pdf
---


## Importing candy data

```{r}

candy_file <- "candy-data.csv"

candy <- read.csv(candy_file,
                    row.names=1)

head(candy)

```

### Q1: How many different candy types are in this dataset?

```{r}

nrow(candy)

```


### Q2: How many fruity candy types are in the dataset?
```{r}

table(candy$fruity)

```
There are 38 fruity candy types.


## What is your favorite candy?

### Q3: What is your favorite candy in the dataset and what is its 'winpercent' value?
```{r}

candy["Kit Kat",]$winpercent

```
My favorite candy in the dataset are Kit Kats, and their win percent value is 76.7686%.

### Q4: What's the winpercent value for Kit Kat?
```{r}

#I didn't notice this was the next question, Kit Kats truly are my favorite 
#for Q3!

candy["Kit Kat",]$winpercent

```

### Q5: What's the winpercent value for Tootsie roll snack bars?
```{r}

candy["Tootsie Roll Snack Bars",]$winpercent

#Or use dplyr package to get the same information
#library(dplyr)
#candy |>
#  filter(rownames(candy)=="Tootsie Roll Snack Bars") |>
#  select(winpercent)

#or
#candy |>
#  filter(rownames(candy) %in% c("Kit Kat",
#                                "Tootsie Roll Snack Bars")) |>
#   select(winpercent)

```



```{r}

#install.packages("skimr")
library("skimr")
skim(candy)

```


### Q6: Is there any variable/column that looks to be on a different scale to the majority of the other columns in the dataset?
The 'winpercent' variable is on a very different scale compared to the rest of the variables, since this is measured as a percent from 0-100, while the majority of the other variable measurements are a binary of 0 or 1. 


### Q7: What do you think a zero and one represent for the candy$chocolate column?
The 0 represents a "no" or "false", and the 1 represents "yes" or "true". If a candy has a 0 in the chocolate variable column, it doesn't have chocolate in it, and if it has a 1 then it does have chocolate.


### Q8: Plot a histogram of winpercent values.
```{r}

hist(candy$winpercent)

library(ggplot2)
ggplot(candy, aes(winpercent)) +
  geom_histogram(bins=10) +
  theme_bw()

```

### Q9: Is the distribution of winpercent values symmetrical?
No, the data shown in the histogram has a positive/right skew with most of the data centered around 30-60% and a tail extending out towards 100%.

### Q10: Is the center of the distribution above or below 50%?
The center is below 50%.


### Q11: On average, is chocolate candy higher or lower ranked than fruit candy?
```{r}

mean(candy$winpercent[as.logical(candy$chocolate)])
mean(candy$winpercent[as.logical(candy$fruity)])

```
Chocolate candy is ranked higher on average than fruit candy. The average win percent for chocolate candy is around 61% while the average win percent for fruit candy is 44%.


### Q12: Is the difference in Q11 statistically significant?
```{r}

chocolate_candy <- as.logical(candy$chocolate)
chocolate_winpercent <- candy[chocolate_candy,]$winpercent

fruit_candy <- as.logical(candy$fruity)
fruit_winpercent <- candy[fruit_candy,]$winpercent

t.test(chocolate_winpercent, fruit_winpercent)

```
According to this t test, there is a significant difference between the mean winpercent values of chocolate and fruity candies. The p value is very low at 2.87E-8.



## Overall candy rankings


### Q13: What are the five least liked candy types in this set?

```{r}


library(dplyr)
candy %>%
  arrange(winpercent) %>%
  head(5)

#I prefer the dplyr method of sorting the dataset because the code makes 
#more logical sense to me as written

```

Nik L Nip, Boston Baked Beans, Chiclets, Super Bubble, and Jawbusters



### Q14: What are the top 5 favorite candy types from this set?
```{r}

candy %>%
  arrange(desc(winpercent)) %>%
  head(5)

```
Reese's PB cups, Reese's miniatures, Twix, Kit Kats, and Snickers


### Q15: Make a barplot of candy ranking based on winpercent values.
```{r}

ggplot(candy) +
  aes(winpercent, rownames(candy)) +
  geom_col()

```


### Q16: Now use the reorder() function to get the bars sorted by wnpercent to clean up the plot a bit.
```{r}

ggplot(candy) +
  aes(winpercent, reorder(rownames(candy), winpercent)) +
  geom_col()

```


```{r}

#Set up a color vector to use for the plot by types of candy
#First make a vector of all black values for each candy.
my_cols = rep("black", nrow(candy))

#Then overwrite the vector to create colors for different candy types
my_cols[as.logical(candy$chocolate)] = "chocolate"
my_cols[as.logical(candy$bar)] = "brown"
my_cols[as.logical(candy$fruity)] = "pink"

#Color my favorite candy in the plot my favorite color
my_cols[rownames(candy) == "Kit Kat"] = "darkgreen"

#Now plot again with these colors
ggplot(candy) + 
  aes(winpercent, reorder(rownames(candy),winpercent)) +
  geom_col(fill=my_cols) 
  #Use fill=mycols (makes the entire bar the desired color) here instead of 
  #col=my_cols (which would only make the outline of the bars the desried color, 
  #no good)

```

### Now, using this plot we can answer:
### Q17: What is the worst ranked chocolate candy?
Sixlets

### Q18: What is the best ranked fruity candy?
Starburst


## Taking a look at pricepercent

```{r}

#Scatterplot of winpercent vs pricepercent with clear labels on each point

library(ggrepel)
ggplot(candy) +
  aes(winpercent, pricepercent, label=rownames(candy)) +
  geom_point(col=my_cols) +
  geom_text_repel (col=my_cols, size=3.3 ,max.overlaps=5)

```

## Q19: Which candy type is the highest ranked in terms of winpercent for the least money?
Based on the plot above, Reese's miniatures are the highest ranked winpercent for the lowest pricepercent.



## Q20: What are the top 5 most expensive candy types in the dataset, and of these, which is the least popular?
```{r}

ord <- order(candy$pricepercent, decreasing=TRUE)
head(candy[ord, c(11,12)], n=5)

```
The top 5 most expensive types are Nik L Nip, Nestle Smarties, Ring pop, Hershey's Krackel, Hershey's Milk chocolate. The least popular of these is Nik L Nip.



### Q21: Make a barplot again with geom_col() this time using pricepercent and then improve this step by step, first ordering the x-axis by value and finally making a so called “dot chat” or “lollipop” chart by swapping geom_col() for geom_point() + geom_segment().
```{r}

#ggplot(candy) +
#  aes(pricepercent, rownames(candy)) +
#  geom_col()

#ggplot(candy) +
#  aes(pricepercent, reorder(rownames(candy), pricepercent)) +
#  geom_col()

ggplot(candy) +
  aes(pricepercent, reorder(rownames(candy), pricepercent)) +
  geom_segment(aes(yend = reorder(rownames(candy), pricepercent), 
                   xend = 0), col="gray40") +
    geom_point()

```


## Exploring the correlation structure

```{r}

library(corrplot)
cij <- cor(candy)
corrplot(cij)

```
### Q22: What two variables are anti-correlated?
Fruity and chocolate are the most anti-correlated variables.

### Q23: What two variables are the most positively correlated?
Chocolate and winpercent are most strongly correlated, meaning chocolate candies are correlated with high winpercent values. 


## Principal Component Analysis

```{r}


pca <- prcomp(candy, scale=TRUE)
summary(pca)

```

```{r}

#Plot PCA score plot of PC1 vs PC2
plot(pca$x[,1], pca$x[,2], col=my_cols, pch=16)

```


```{r}

#To make a ggplot plot, first make a new data frame with our PCA results 
  #(PC1-3) and candy data together
my_data <- cbind(candy, pca$x[,1:3])

p <- ggplot(my_data) +
      aes(PC1, PC2,
          size=winpercent/100,
          text=rownames(my_data),
          label=rownames(my_data)) +
      geom_point(col=my_cols)
p

```


```{r}

#Use ggrepel to label with non-overlapping names 

p + geom_text_repel(size=3.3, col=my_cols, max.overlaps = 7)  + 
  theme(legend.position = "none") +
  labs(title="Halloween Candy PCA Space",
       subtitle="Colored by type: chocolate bar (dark brown), chocolate other 
       (light brown), fruity (pink), other (black)",
       caption="Data from 538")

```

```{r}

#Use plotly to generate an interactive plot!
#install.packages("plotly")
#library(plotly)
#ggplotly(p)

```


```{r}

#Bar plot to visualize PC1 contributions 
par(mar=c(8,4,2,2))
barplot(pca$rotation[,1], las=2, ylab="PC1 Contribution")

```
### Q24: What original variables are picked up strongly by PC1 in the positive direction? Do these make sense?
The fruity, hard, and pluribus variables are strongly picked up by PC1 in the positive direction. These make sense because looking at the PC1 vs PC2 scatterplot, the candies on the rightmost part of the x axis are mainly hard fruity candies that come in multi pack packaging. The candies on the leftmost part of the x axis are mainly chocolate bars. 


